---
title: "Tornado Data EDA"
author: "JMG"
format: html
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggthemes)

theme_set(theme_minimal(base_size = 12))
```



## The Data

We obtained the raw data from [here(https://www.spc.noaa.gov/wcm/data/1950-2022_actual_tornadoes.csv)]. It contains observations about tornado events from 1950-2022. The data is downloaded in the R script "get_tornados_data.R". The cleaned data is saved in the data folder as "tornados.csv"

Let's load the data:

```{r}
tornados <- read_csv("data/tornados.csv")
```

Check what we read in

```{r}
glimpse(tornados)
```

This data contains 68,693 rows and 27 columns (variables).

## Exploratory Data Analysis

Initial examination of entire data 
```{r}
skimr::skim(tornados)
```

Note: There are a lot of missing values for "loss" and a fair number for "mag" as well. 

```{r}
tornados %>%
  ggplot(aes(x = st)) + 
  geom_bar() + 
  coord_flip()
```

Note that there are 50 states and 3 territories included.

```{r}
tornados %>%
  ggplot(aes(x=slon,y=slat)) + 
  geom_point(alpha=0.1)
```

```{r}
states_df <- map_data("state")
```

```{r}
ggplot() + 
  geom_polygon(data=states_df,
               mapping = aes(long,lat,group=group),
               fill="white",color="darkgray")
```

Build our map data and filter out AK, HI, PR, and VI

```{r}
tornados_49 <- tornados %>%
  filter(st != "AK", st != "HI",st!= "PR", st!= "VI")
```

```{r}
tornados_49 %>%
  ggplot(aes(x = st)) + 
  geom_bar() + 
  coord_flip()
```

From which we see the states with the most tornados:

```{r}
#| label: fig-main-states
#| fig-cap: The continental 48 U.S. States and Washington D.C. together with the frequency of tornados observed in each.

tornados_49 %>%
  group_by(st) %>%
  summarise(per_state = n()) %>%
  arrange(desc(per_state)) %>%
  ggplot(aes(x=reorder(st,per_state),y=per_state)) + 
  geom_bar(stat = "identity") + 
  coord_flip()
```

Filter out (that is, remove) those observations where the ending latitude or longitude is 0.

```{r}
tornados_49 <- tornados_49 %>%
  filter(elat > 0 & elon < 0)
```

```{r}
tornados_49 %>%
  ggplot() + 
  geom_polygon(data=states_df,
               mapping = aes(long,lat,group=group),
               fill="white",color="darkgray") + 
  geom_segment(aes(x=slon,y=slat,xend=elon,yend=elat), alpha=0.2)
```


Remove observations from the West. 

```{r}
tornados_49 <- tornados_49 %>%
  filter(slon >= -105)

states_df <- states_df %>%
  filter(long >= -105)
```


```{r}
#| Label: fig-tornadoes-map
#| fig-cap: Map showing location and distance of travel for tornadoes observed in data.
tornados_49 %>%
  ggplot() + 
  geom_polygon(data=states_df,
               mapping = aes(long,lat,group=group),
               fill="white",color="darkgray") + 
  geom_segment(aes(x=slon,y=slat,xend=elon,yend=elat),alpha=0.5,color="darkgreen")
```

Frequency of tornados by year.
```{r}
tornados_49 %>%
  ggplot(aes(x= yr)) + 
  geom_bar()
```

Tornados by month. 

```{r}
tornados_49 %>%
  ggplot(aes(x=mo)) + 
  geom_bar()
```

```{r}
tornados_49 %>% 
  filter(fat > 0) %>%
  ggplot(aes(x=log10(fat))) +
  geom_histogram()
```

Comparing estimated versus measured magnitudes 

```{r}
tornados_49%>%
  mutate(mag=factor(mag)) %>%
  ggplot(aes(x=mag,fill=fc)) +
  geom_bar(position=position_dodge(preserve="single")) + 
  scale_y_log10()+
  labs(x="Magnitude", y="Count", fill="Estimated")
```

```{r}
tornados_49%>%
  filter(loss >0) %>%
  ggplot(aes(x=log10(loss))) +
  geom_histogram
```

```{r}
tornados_49 %>%
  mutate(ns=factor(ns)) %>%
  ggplot(aes(x=ns)) +
  geom_bar() + 
  scale_y_log10()
```

```{r}
```


```{r}
tornados_49%>%
  ggplot(aes(x=len)) +
  geom_histogram() +
  scale_x_log10()
```


```{r}
tornados_49%>%
  ggplot(aes(x=wid)) +
  geom_histogram() +
  scale_x_log10()
```

```{r}
tornados_49 %>%
  mutate(mag=factor(mag)) %>%
  ggplot(aes(x=mag, y=len)) +
  geom_boxplot()
```

```{r}
tornados_49 %>%
  filter(inj>0) %>%
  mutate(mag=factor(mag)) %>%
  ggplot(aes(x=mag, y=inj)) +
  geom_boxplot() 

```